---
title: " Analyze Daily Step Activjty from an Activity Monitor"
author: "Hadia Akbar"
output: 
  html_document: 
    keep_md: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document aims to analyze daily activity pattern recorded using an activity monitor.

##1-Read and process the data
```{r}
library(readr)
data <- read_csv("activity.csv", col_types = cols(date = col_date(format = "%m/%d/%Y")))
```

##2-The total number of steps taken each day
```{r}
sum <- data.frame(tapply(data$steps,data$date,sum,na.rm=TRUE))
date <- data.frame(date=as.Date(row.names(sum)))
names(sum)[[1]]<-"TotalSteps"
```
Histogram for steps taken each day
```{r}
library(ggplot2)
ggplot(sum,aes(TotalSteps))+
  geom_histogram(fill="blue")+
  labs(title="Steps Taken Each Day",x=" Number of Steps", 
       y ="Frequency")+  
  theme_classic()
```

##3.	Mean and median number of steps taken each day
Mean
```{r}
mean(sum$TotalSteps)
```
Median
```{r}
median(sum$TotalSteps)
```

##4.  Time series plot of the average number of steps taken
```{r}
#Calculate average for number of steps taken in each interval
Ave <- data.frame(AveSteps=round(tapply(data$steps, data$date,mean),2), na.rm=TRUE)
Ave$date <- as.Date(row.names(Ave))
rownames(Ave) <- NULL
#remove na values form the data
Ave <- na.omit(Ave)
```
Time Series plot.
```{r}
ggplot(Ave, aes(date, AveSteps))+
  geom_line(color="#00AFBB", group=1)+
  labs(title="Avergae Steps Each Day",x="Date",y="Average Steps")+
  theme_classic()
```

  ## 5 minute interval that had the maximum number of steps
```{r}
#subset the data to get the interval with the maximum number of steps
subset(data, steps==max(data$steps))
```

##6	Imputing missing data
```{r}
#Number of rows with NA
nrow(data[is.na(data$steps),])
```
There are many ways to impute missing data, hereI choose to fill NA values with mean of data
```{r}
#Create a new dataset that is equal to the original dataset but with the missing data filled in.
StepsPerInterval <- round(tapply(data$steps, data$interval, mean, na.rm = TRUE),0)

#Split data data by interval
data.split <- split(data, data$interval)
#Create a for loop and fill in missing data for each interval
for(i in 1:length(data.split)){
  data.split[[i]]$steps[is.na(data.split[[i]]$steps)] <- StepsPerInterval[i]
  }
#Bind the split data with filled 'na' to get a new data set
newdata <- do.call("rbind", data.split)
newdata <- newdata[order(newdata$date) ,]
```

##7	Histogram of the total number of steps taken each day after missing values are imputed
```{r}
ggplot(newdata,aes(steps))+
  geom_histogram(fill="blue")+
  labs(title="Steps Taken Each Day-Imputed Data",
       x=" Number of Steps", 
       y ="Frequency")+
  theme_classic()

sum.imputed <- data.frame(tapply(newdata$steps, newdata$date,sum, na.rm=TRUE))
date <- data.frame(date=as.Date(row.names(sum.imputed)))
names(sum.imputed)[[1]]<-"TotalSteps"
```
Calculate and report the mean and median total number of steps taken per day. 
Mean
```{r}
mean(sum.imputed$TotalSteps)
```
Median
```{r}
median(sum.imputed$TotalSteps)
```

##8-	Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r plot, include=FALSE}

library(reshape2)
library(plyr)
#Create a new variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend.
newdata$day <- ifelse(weekdays(as.Date(newdata$date))== "Saturday" | weekdays(as.Date(newdata$date)) == "Sunday", "weekend", "weekday")

#Calculate average steps by intervals on weekends and weekdays
AveBycategory <- ddply(newdata, .(interval, day), summarize, Avg = mean(steps))

#weekend plot
we<- ggplot(AveBycategory[AveBycategory$day == "weekend",], 
       aes(interval, Avg))+
  geom_line(color = "#00AFBB", size=1)+
  labs(title="Weekend",x="Interval",y="Average Steps")+
  theme_classic()

#weekday plot
wd <- ggplot(AveBycategory[AveBycategory$day == "weekday",], 
              aes(interval, Avg))+
  geom_line(color = "#00AFBB", size=1)+
  labs(title="Weekday",x="Interval",y="Average Steps")+
  theme_classic()
```

```{r plot2, results="hide"}
library("gridExtra")
#bring together the weekend and weekday plots
f<- grid.arrange(we, wd, ncol = 1, nrow = 2)  
library(ggpubr)
annotate_figure(f, top = text_grob("Average steps as per day category",
                          face = "bold", size = 14))
```
